The markings of the trade: exploring the patterns of olive oil production in Roman Baetica

### 2018-06-07 Paper/jaccardbaetica.tex
modify some error
review english

\begin{abstract}

@@

CAMBIAR

@@

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors have an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. As a result, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places.

@@

%This paper explore--- resumir en dos líneas el objeto de estudio y definir que ha sido usado para ello dispersion of culture
This paper aims to understand the production dynamics in relation to a specific area within Roman Empire. Specifically, our work pretend to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to detect the pattern of olive oil production that link amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production centres. To do this, an Ecology approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}.

@@

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were created to support the high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped thought the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in graffities with different information. However, graffities and \textit{tituli picti} have not enough analysed due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae. During almost three centuries, stamps were used to mark amphorae with a different chronological frequency \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criteria. Stamps was mostly formed for a code of three letters. These letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

@@

Here, therefore, this study proposes an approach to explore the effect of Baetican olive oil production by computing of dissimilarity index. The assumption of this analysis is that closer workshops concentrate similar amphoric stamps in a specific area than the farthest workshops. By contrast, if we found similar amphoric stamps in different areas between workshops then the correlation between spatial distance would not be valued.

@@

Our case study examines the relation between the distribution of amphoric stamps and the workshops. The workshops were situated in different locations in Baetica province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop})

@@

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production being difficult to specify an accurate chronology. This could be due to two reasons. First, most of workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible. Second, this amphora type was produced during almost three centuries with apparently few changes so they presented a long chronology. (PODRÍA CITARME AQUÍ MUAHA)

We studied a dataset of 3787 stamps collected from different Dressel 20 amphora workshops in Baetica province. The stamp database was compiled by CEIPAC database \citep{remesal_centro_2015} (see CEIPAC database). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filter a total sample of 987 stamps composed by 131 different stamps from 81 workshops.

@@

The distribution of amphora stamps in different conventus can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly workshops show a homogeneity on frequency of stamps except La Catria and Arva (both show 29 different stamps) that show a big amount of different stamps. According to some researchers, those workshops became in most important centres of amphora production although it could have been more intensely prospected then others. \citep{arva_1997}

@@

The approach proposed here is based on the idea of measuring the similarity between amphora workshops by quantifying similar stamps. A measure of dissimilarity has been chosen to analyse the dataset. We use the statistical technique Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was performed to measure the overlap between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}.

@@

where $N_{a}$ is the total number of stamps in workshop A; $N_{b}$ is the total number of stamps in workshop B; $a_{i}$ is the number of different stamps for workshop A and $b_{i}$ is the number of different stamps for workhop B.

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. This dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, no groups of similar stamps were found in the cluster. The majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, as for instance Picachos, Cerro de los Pesebres and El Castillejo.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendro}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different amphora workshops in Baetica area. Colors are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{dendro}
\end{figure}

@@

In our analysis, no strong relation between stamps and geographic distance were detected in amphora workshops. Indeed, the analysis suggests that there is not connection between stamps and the same amphora workshops, excluding certain exceptions when nearby workshops share the same amphoric stamps. Consequently the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. In any case, our results show that most similar stamps were detected in the same \textit{conventus} area. These stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and area.

The hypothesis about groups of amphora workshops sharing the same stamps seems do not match with the results of the analysis even though there are similar stamps in closer workshops. Rather, it seems that each workshop were organized independently with different stamps. Those stamps detected in closer workshops do not move from other farthest workshops. In other words, the stamps tend to keep in the same area and different stamps were located in a same amphora workshops. 
This could be defined by several factors. Firstly, each workshop had a different organization involved to the use of stamps and they were not used in other workshops. Secondly, stamps similarity in closer workshops could be linked to a spatial pattern. It is more probably than closer workshops tend to share more traits than farthest. While the role of river was significant for the distribution of amphorae in consumption places, river connection amongst workshops was not relevant for the distribution of stamps. Finally, the distribution of stamps could have showed some research bias. In some cases, workshops have been catalogued with different names despite belonging to the same workshops or being closer between each other. Additionally, most of the workshops were not widely excavated.

%soltar aquí las hipótesis

The results of the case study could be interpreted due to several reason according to previous hypotheses. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop (CITAR). This hypotheses would explain the fact that we do not find similar stamps in different workshops, however, we found different amphora stamps in a same workshops that they would be barely difficult to assign different owners. Neither a kind of quality distinction does seem to be used to specify the quality or value of the product \citep{callender}. On the other hand, it can be interpreted somehow a batch systematic organization. Considering that Dressel 20 was not marked in several cases, some authors point out that potters marked amphorae to prepare and distribute the commodity to be shipped \citep{berni_millet_epigrafianforica_2008}. This method would be used as an identifier to count the number of amphorae of a branch (CITAR). This method could also have served to identify different groups of potter workers working in the same amphora workshop. Potters could have marked the amphorae to distinguish  different groups working in parallel \citep{li_crossbows_2014}. This it could be explain wherefore we found different stamps in a same workshop. 
In any case, we do not have enough archaeological evidence that can validate the interpretations presented here and our results are certainly valid only with the context of our case study.

As summary, this method presented here provides a potential tool to understand mechanisms of production based on the similarity of artefacts. This method have identified differences in the case of the amphoric production within Roman Empire. Accordingly the results have highlighted to interpreter the complex economical processes based on the archaeological evidence.

@@

\section{Acknowledgements}

The research was funded by European Research Council Advanced Grant EPNet (340828). We are grateful to Simon Carrignon, Juan Moros and Ignacio Morer for their useful suggestions.  
All data has been analysed and conducted in R program version 3.2.4, using the packages \textit{vegan} \citep{oksanen_vegan_2007}, \textit{ggplot2} \citep{ggplot2:_2016}. Source and code are available

>>>>

%cambiar quizás
\begin{abstract}

@@

@@

%primer párrafo más general---- especifico (¿dónde se ha aplicado?)
Material culture is one of the most frequent indicator of trade in the archaeological record. In archaeology, they allow us to highlight a part of mechanism of production and distribution of goods along the Mediterranean \citep{bevan_mediterranean_2014}. Particularly, the spread of these factors had an important impact during the Roman Age, when the progressive exploitation of communication networks allowed a major interaction between communities \citep{orengo_seeds_2016}. As a result, an important mechanism of production control under Roman government was created spreading by different areas attending to the richness of different places.

@@

%This paper explore--- resumir en dos líneas el objeto de estudio y definir que ha sido usado para ello dispersion of culture
This paper aims to understand the production dynamics in relation to a specific area within Roman Empire. Specifically, our work pretends to detect microeconomic processes focused on a commercial product from a specific province \citep{isaksen_network_2006}. We want to detect the pattern of olive oil production that link amphora workshops and amphoric stamps used to mark them. We focus here on exploring the economic relation between stamps and amphora production centres. To do this, an Ecology approach has been used to analyse the dispersion of stamps between amphora workshops \citep{rubio-campillo_ecology_2018}.

@@

The production and distribution of olive oil in this ancient province were growing exponentially during almost three centuries \citep{remesal_concierto}. As a result, hundred of amphora workshops were created to support the high demand of Roman Empire. These workshops were located along the rivers Guadalquivir and Genil, supplying from the riverine connectivity to the Mediterranean and Atlantic routes \citep{garcia_vargas_enrique_formal_2010}.

The majority of olive oil amphorae produced in this province and shipped through the provinces belong to Dressel 20 type \citep{dressel,martin-kilcher_romischen_1994}. This amphora is commonly associated with the transportation of Baetican olive oil during the Roman Empire \citep{berni_millet_epigrafianforica_2008}. Most Dressel 20 were marked in stamps, inked in \textit{tituli picti} and incise in \textit{graffiti} with different information. However, inscriptions and \textit{tituli picti} have not enough analysed by researchers due to the fragmentation of the material or the shortage of samples \citep{aguilera_evolucion_2007,rovira_guardiola_grafitos_2007}. By contrast, stamps are the most studied in this type of amphorae. During almost three centuries, stamps were used to mark amphorae with a different chronological frequency \citep{remesal_sellar_2016}. Frequently, they were marked mainly in handles but rarely in rims and body \citep{millet_anforas_1998}. 
The information of the stamps is shown in different forms and letter content and it seems that there was not a unique criterion. Stamps was mostly formed for a code of three letters. These letters can appear in a abbreviated form or complete and they are known as \textit{Tria Nomina} \citep{berni_millet_amphora_1996}.

@@

Here, therefore, this study proposes an evolutionary approach to explore the effect of Baetican olive oil production by computing of dissimilarity index. The assumption of this analysis is that closer workshops concentrate similar amphoric stamps in a specific area than the farthest workshops. By contrast, if we found similar amphoric stamps in different areas between workshops then the correlation between spatial distance would not be valued.

@@

Our case study examines the relation between the distribution of amphoric stamps and the workshops. The amphora workshops were situated in different locations in Baetica province, along the river Guadalquivir and its tributary Genil in order to detect similarities between stamps from workshops and spatial distance (see Fig.\ref{workshop})

@@

The chronology in the workshops is widely diverse from the first to the third centuries AD \citep{millet_anforas_1998,rodriguez_baetican_1998,chic2005comercio}. Some stamps show a more specific chronology while the majority of them display a large activity of production being difficult to specify an accurate chronology. This could be due to two reasons. Firstly, most of the workshops were partially excavated and focused on archaeological surveys in order to collect the maximum stamps as possible. Secondly, this amphora type was produced during almost three centuries with apparently few changes so they presented a long chronology. (podría citar nuestro paper por toda la face).

We studied a dataset of 3787 stamps collected from different Dressel 20 amphora workshops in Baetica province. The stamp database was compiled by CEIPAC database \citep{remesal_centro_2015} (see CEIPAC database here \url{http://romanopendata.eu}). However, approximately the 70 \% of stamps cannot be tested due to fragmentation or incomplete information. Consequently, we discard integrate the fragmented stamps in our dataset. We finally filter a total sample of 987 stamps composed by 131 different stamps from 81 workshops.

@@

The distribution of amphora stamps in different \textit{conventus} can be seen in Fig. \ref{frequency}. The majority of stamps found are concentrated in \textit{Hispalis} with 574 stamps while \textit{Corduba} and \textit{Astigi} with 267 and 146 stamps, respectively. Mostly workshops show a homogeneity on frequency of stamps except La Catria and Arva (both show 29 different stamps) that show a big amount of different stamps. According to some researchers, those workshops became in most important centres of amphora production although they could have been more intensely prospected then others. \citep{arva_1997}

@@

The approach proposed here is based on the idea of measuring the similarity between amphora workshops by quantifying similar stamps. A measure of dissimilarity has been chosen to analyse the dataset. We use the statistical technique Morisita-Horn index \citep{morisita_measuring_1959, horn_measurement_1966}. This method was performed to measure the dissimilarity between different samples of sets. Generally, it describes the dissimilarity between the system of two communities based on the idea of inverse correlation between diversity and species \citep{magurran_why_1988}.

@@

where $N_{a}$ is the total number of stamps in workshop A; $N_{b}$ is the total number of stamps in workshop B; $a_{i}$ is the number of different stamps for workshop A and $b_{i}$ is the number of different stamps for workshop B.

@@

The analysis shows that amphoric stamps could be correlated with the spatial distance. The correlation coefficients range from a minimum to a maximum. The dendrogram shown in Fig. \ref{dendro} was obtained with Morisita-Horn index. The dendrogram suggests that amphora workshops used different stamps for their production system. Nearby workshops show a similarity on the stamps while most of them seem to display different stamps roles. Additionally, no groups of similar stamps were found in the cluster. The majority of stamp grouping was composed by no more than three workshops. Indeed workshops that shared more similar amphoric stamps belonged to the same \textit{conventus} area, such as Picachos, Cerro de los Pesebres and El Castillejo.

\begin{figure}[htp]
	\centering
\includegraphics[width=\linewidth]{figs/dendro}
\caption{Dendrogram obtained by Morisita-Horn algorithm of different amphora workshops in Baetica area. Colours are represented by areas divided into Hispalis (red), Astigi (green) and Corduba (blue)}
\label{dendro}
\end{figure}

@@

In our analysis, no strong relation between stamps and spatial distance were detected in amphora workshops. Indeed, the analysis suggests that there is not connection between stamps and the same amphora workshops, excluding certain exceptions when nearby workshops share the same amphoric stamps. Consequently, the majority of stamps are located in different amphora workshops and only similar stamps between closer amphora workshops were found. In any case, our results show that most similar stamps were detected in the same \textit{conventus} area. These stamps tend to share the same area of production but there is not a general relation between groups of amphora workshops and area.

The hypothesis about groups of amphora workshops sharing the same stamps seems do not match with the results of the analysis even though there are similar stamps in closer workshops. Rather, it seems that each workshop were organized independently with different stamps. Those stamps detected in closer workshops do not move from other farthest workshops. In other words, the stamps tend to keep in the same area and different stamps were located in a same amphora workshops.

This could be defined by several factors. First, each workshop had a different organization involved to the use of stamps and they were not used in other workshops. Second, stamp similarity in closer workshops could be linked to a spatial pattern. It is more probably than closer workshops tend to share more traits than farthest. While the role of river was significant for the distribution of amphorae in consumption places, river connection amongst workshops do not seem show a relevancy for the distribution of stamps. Finally, the distribution of stamps could have showed some research bias. In some cases, workshops have been catalogued with different names despite belonging to the same workshops or being closer between each other. Additionally, most of the workshops were not widely excavated.

%soltar aquí las hipótesis repito aquí several reasons no repetir

The results of the case study could be interpreted due to several reason according to previous hypotheses. On the one hand, the use of these amphoric stamps could have been exclusively running by the owner or owners of the workshop to distinguish the amphora workshop (CITAR). This hypothesis would explain the fact that we do not find similar stamps in different workshops, however, we found different amphora stamps in the same workshops that they would be barely difficult to assign different owners. Neither a  quality distinction does seem to be used to specify the value of the product \citep{callender}. On the other hand, it can be interpreted somehow a batch systematic organization. Considering that Dressel 20 was not marked in several cases, some authors point out that potters marked amphorae to prepare and distribute the commodity to be shipped \citep{berni_millet_epigrafianforica_2008}. This method would be used as an identifier to count the number of amphorae of a branch (CITAR). This method could also have served to identify different groups of potter workers working in the same amphora workshop. Potters could have marked the amphorae to distinguish different groups working in parallel \citep{li_crossbows_2014}. This it could be explain wherefore we found different stamps in a same workshop. 
In any case, we do not have enough archaeological evidence that can validate the interpretations presented here and our results are certainly valid only with the context of our case study.

As summary, this method presented here provides a potential tool to understand mechanisms of production based on the similarity of artefacts. This method have identified differences in the case of the amphoric production within Roman Empire. Accordingly the results have highlighted the interpretation of the complex economic processes based on the archaeological evidence.

@@

\section{Acknowledgements}
%completar 
The research was funded by European Research Council Advanced Grant EPNet (340828). We are grateful to Simon Carrignon, Juan Moros and Ignacio Morer for their useful suggestions.  
All data has been analysed and conducted in R program version 3.2.4, using the packages \textit{vegan} \citep{oksanen_vegan_2007}, \textit{ggplot2} \citep{ggplot2:_2016}. Source and code are available
